Reasoning Graphically about Quantum Computation

### 2008-02-28 papers/reasoning-graphically/paper.tex
fixed typo
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1385 f8f375fb-4f10-0410-8d47-b6fbdc7ca053
some images, intro for var-nodes and reasoning section work
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1386 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

\begin{figure}{t}
  We make the convention that domain of an arrow is at the top of the
  picture, and its codomain is at the bottom.
  \centering
  \[
  \begin{array}{lllll}
      \id{A} \otimes \id{B^*} = \id{A \otimes B^*} = PIC 
      &\qquad&
      d_A = PIC  
      &\qquad& 
      f = PIC \\\\
      \name{f} = PIC 
      & \qquad &
      \coname{f} = PIC 
      &\qquad &
      f^* = PIC
  \end{array}
  \]
  \caption{Compact Closed Structure as Graphs}
  \label{fig:comcl-graphs}
\end{figure}

@@

Variable nodes generalise the concept of a boundary nodes. Unlike
boundary-nodes in the concrete graph, they are not classified into
domain and co-domain. The intuition of a variable node is that it can
be replaced by any subgraph of the right type. Thus, unlike nodes in
conrete a graph, variable nodes do not define an ordering on the edges
that are incident. Instead, the edges that touch a variable node are
simply a set. The type of a variable nodes is thus defined as a
multiset rather than a tensor. Any tensor type can be lifted to a
multiset type in the obvious manner, which we denote by the operation
{\tt mset\_of\_tensor}. For example, {\tt mset\_of\_tensor($a \tensor
  b \tensor a$) = $\{(a,2), (b,1)\}$}. We say that a tensor-type $T$
is an instance of a multiset type $M$, written $T \in M$, when $M
\subseteq \mathtt{mset\_of\_tensor}(T)$. The intuition here is that
when a variable node is instantiated to some other node, it's type
will have at least the currently attached edges.

@@

\subsection{Matching and Open Embedding}

\begin{definition}
  An \emph{open embedding} from $G$ to $H$ a strict embedding from the
  non-variable subgraph of $G$ into the non-variable subgraph of $H$
  and a mapping from the variable nodes in $G$ ...
\end{definition}

@@

The formalism we have presented for graph patterns provides a
\emph{meta-level} framework for reasoning in some model of concrete
graphs.

An \emph{object-level} graph formalism is defined by some
model of a compact closed category.

This includes a particular
representation for data at the nodes and edges as well as
node-matching and edge-matching algorithms. The object-level
definition also defines a set of axiomatic equational rules.

Such a theorem prover starts with a system of equations between
graphs. These are not true of the graphs abstractly, but are proved of
the particular model of concrete graphs being considered.

@@

\subsection{Rules}

>>>>

\begin{figure}[t]
  We make the convention that domain of an arrow is at the top of the
  picture, and its codomain is at the bottom.
  \centering
  \[
  \begin{array}{lllll}
      \id{A} \otimes \id{B^*} = \id{A \otimes B^*} = PIC 
      &\qquad&
      d_A = PIC  
      &\qquad& 
      f = PIC \\\\
      \name{f} = PIC 
      & \qquad &
      \coname{f} = PIC 
      &\qquad &
      f^* = PIC
  \end{array}
  \]
  \caption{Compact Closed Structure as Graphs}
  \label{fig:comcl-graphs}
\end{figure}

@@

In the concrete representation, the following graphs represent
different computations:

\begin{center}
\inlinegraphic{2.5em}{compact1} \quad\quad
\inlinegraphic{2.5em}{compact3} \quad\quad
\inlinegraphic{2.5em}{compact5}
\end{center}

\noindent However, composition with semicircle line(s) on the domain
or co-domain allows a proof involving one of the above to easily
derive a corresponding one for the other two. Proving any of the
following allows a trivial derivation of the others:

\begin{center}
\inlinegraphic{2.5em}{compact1} $=$
\inlinegraphic{1.3em}{compact2} $\Leftrightarrow$
\inlinegraphic{2.5em}{compact3} $=$ \inlinegraphic{1.3em}{compact4}
$\Leftrightarrow$ \inlinegraphic{2.5em}{compact5} $=$
\inlinegraphic{2.5em}{comonoid-unit4}
\end{center}

\noindent This motivates a representation that has only one
representative for such equations. Such a representation abstracts
over the boundary nodes membership in the domain and co-domain. In
this section we formalise such a notion and refer to the generalised
boundary nodes as variable-nodes.

The intuition of a variable node is that it can be replaced by any
subgraph of the right type. Thus, unlike nodes in conrete a graph,
variable nodes do not define an ordering on the edges that are
incident. Instead, the edges that touch a variable node are simply a
set. The type of a variable nodes is thus defined as a multiset rather
than a tensor. Any tensor type can be lifted to a multiset type in the
obvious manner, which we denote by the operation {\tt
  mset\_of\_tensor}. For example, {\tt mset\_of\_tensor($a \tensor b
  \tensor a$) = $\{(a,2), (b,1)\}$}. We say that a tensor-type $T$ is
an instance of a multiset type $M$, written $T \in M$, when $M
\subseteq \mathtt{mset\_of\_tensor}(T)$. The intuition here is that
when a variable node is instantiated to some other node, it's type
will have at least the currently attached edges.

@@

\subsection{Matching as an Open Embedding}

To formalise the semantics of variable-node graphs, we define
matching, which captures the intuitive idea with

\begin{definition}
  Matching is charcterised by an \emph{open embedding} which is strict
  for the non-variable nodes.
\end{definition}

@@

In this section we show how the formalism we have presented for graph
patterns can be used to provide a \emph{meta-level} framework for
reasoning about a given model of a compact closed
categories. Following the terminology of logical frameworks such as
Isabelle~\cite{isabelle}, we call the specification provided by the
underlying model an \emph{object-level} graph formalism. Each
object-level formalism defines a set of rules which are treated as the
axioms for the system. The object-level implementation also defines
the data at the nodes and edges as well as corresponding data-matching
behaviour.

These are not true of the graphs abstractly, but are proved of the
particular model of concrete graphs being considered.

\subsection{Substitution with Equational Rules}

A rule in our framework is a pair of pattern graph that represent the
left and right hand sides of an equation. Substituon with a rule then
involves replacing a subgraph that matches the left hand side with the
right hand side. To do this there has to be a isomorphism between
variable nodes in the left and right hand subgraphs. More concretely,
there must be a unique naming of the variable nodes in the left hand
side such that each name occurs in a unique variable node on the right
hand side. Given a matching on the left hand side, the target subgraph
can be replaced with right hand side connected by the corresponding
variable nodes. This provides rewriting for pattern graphs.

@@

\subsection{Implications}
